---
title: "cancensus"
subtitle: "an R interface to Canadian Census data"
author: "Dmitry Shkolnik"
date: "2017/10/03"
output:
  xaringan::moon_reader:
    css: ["default", "ubuntu-mono.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
## Language Diversity in Metro Vancouver

```{r language_code slide, message=FALSE, warning=FALSE, include=FALSE, results='hide'}
library(cancensus)
library(sf)
library(ggplot2)

# Select Census dataset
dataset <- "CA16"

# Get Language Parent Vector
language_total <- search_census_vectors("Language spoken most often", dataset) %>% 
  filter(vector == "v_CA16_1355") 

# Select all leaf nodes of this vector. The parameter TRUE returns only the finite leaves among child nodes. 
language_children <- language_total %>%
  child_census_vectors(TRUE) 

# We'll need the aggregated total for our calculations so let's append them together
language_vectors <- bind_rows(language_total, language_children) %>%
  pull(vector)

# Get regions list for Van CMA
regions_list_van <- list_census_regions(dataset) %>% 
  filter(level=="CMA", name == "Vancouver") %>% 
  as_census_region_list

langs_van <- get_census(dataset, level = "DA", regions = regions_list_van , vectors = language_vectors, geo_format = "sf", labels = "short")

# Function to compute diversity index
ldi_calc <- function(df) {
  tidy_langs <- df %>%
  rename(`Language Total` = v_CA16_1355) %>%
  mutate(v_CA16_1364 = v_CA16_1364 + v_CA16_2153 + v_CA16_2150 + v_CA16_2159,
         v_CA16_1367 = v_CA16_1367 + v_CA16_2156 + v_CA16_2150 + v_CA16_2159) %>%
  select(-v_CA16_2153, -v_CA16_2150, -v_CA16_2159, -v_CA16_2156) %>%
  tidyr::gather(key = language, value = lang_count, v_CA16_1364:v_CA16_1937) %>%
  mutate(ldi_frac = ifelse(lang_count <= `Language Total`, lang_count/`Language Total`, 1)) %>%
  group_by(GeoUID) %>%
  mutate(ldi = 1 - sum((ldi_frac)^2)) %>%
  ungroup() %>%
  select(-language, -lang_count, -ldi_frac) %>%
  distinct()
  return(tidy_langs)
}

# Apply function over data
van_ldi <- ldi_calc(langs_van) %>%
  mutate(`Region Name` = as.factor(gsub(" \\(.*\\)","",`Region Name`)))

# Theme for maps
map_theme <- theme_void() + 
  theme(plot.title=element_text(face="bold", hjust = 0.5)) + 
  theme(plot.subtitle=element_text(hjust = 0.5)) + 
  theme(plot.caption=element_text(size=8, margin=margin(t=10), hjust = 0.95)) +
  theme(legend.position = "bottom")

# Generate maps
mv1 <- ggplot(van_ldi) + geom_sf(aes(fill = ldi), colour = NA) + scale_fill_viridis_c("Language Diversity Index",option = "magma") +
  map_theme + 
  guides(fill = guide_colourbar(title.position="top"))
  #labs(title = "Language Diversity Index",
  #     subtitle = "Vancouver CMA DA-level data",
  #     caption = "Dmitry Shkolnik @dshkol\nCensus")
```
```{r show_map, echo=FALSE, fig.height=12, fig.width=18, message=FALSE, dpi = 300,  warning=FALSE, fig.align = 'center'}
mv1
```
---
## *cancensus*: An R interface for Canadian Census data that is:

- Reproducible
- Programmatic
- Spatial
- Tidy
- Analysis ready

---
## *cancensus*: An R interface for Canadian Census data that is:

- Reproducible
  - Code for methods, data, and approaches
  - As transparent as you want it to be
- Programmatic
- Spatial
- Tidy
- Analysis ready

---
## *cancensus*: An R interface for Canadian Census data that is:

- Reproducible
- Programmatic
  - Reduce duplication of work
  - Version control for research
- Spatial
- Tidy
- Analysis ready

---
## *cancensus*: An R interface for Canadian Census data that is:

- Reproducible
- Programmatic
- Spatial
  - Binds geographic data to Census data
  - Simple GIS 
- Tidy
- Analysis ready

---
## *cancensus*: An R interface for Canadian Census data that is:

- Reproducible
- Programmatic
- Spatial
- Tidy
  - Follows [tidy](http://vita.had.co.nz/papers/tidy-data.pdf) principles
  - Fits within the R _tidyverse_
- Analysis ready

---
## *cancensus*: An R interface for Canadian Census data that is:

- Reproducible
- Programmatic
- Spatial
- Tidy
- Analysis ready
  - Minimal data prep needed
  
---
## Why Cancensus?

The current Statistics Canada dissemination model wasn't quite cutting it for some people. 

Programmatic API access to Census data has been promised for a long time but we got tired of waiting...


...so we built it ourselves. 

---
## Under the hood

*cancensus* is an open-source library that anyone can install for their R client. It's not yet on CRAN* but you can install the development version from Github.

```{r, eval=FALSE, message=FALSE, warning=FALSE, include=TRUE}
devtools::install_github("mountainmath/cancensus")
```

*cancensus* is an API to Statistics Canada Census data cloned and rehosted via [censusmapper.ca](https://censusmapper.ca). While use of this package requires an API key to manage bandwidth and usage, anyone can register for a free key on censusmapper.ca. 

Once you have downloaded *cancensus* and received your API key, just set your environment up the first time you use it by running:

```{r, eval=FALSE, message=FALSE, warning=FALSE, include=TRUE}
Sys.setenv(CM_API_KEY = "YOUR_KEY")
```
.footnote[[*] Soon. The truth is we got distracted making things with cancensus instead of finishing it.]
---
## A *cancensus* query in detail

```{r, eval=FALSE, message=FALSE, warning=FALSE, include=TRUE}
example <- get_census(dataset, level, regions, 
                      vectors, geo_format, labels, 
                      use_cache)
```
- `dataset`: select Census 2006, 2011, or 2016 data
- `level`: Geographic aggregation level (C, P, CD, CSD, CT, DA)
- `regions`: Census region for your data
- `vectors`: Census variable vectors
- `geo_format`: data only, or data + spatial in `sp` or `sf` formats
- `labels`: Detailed vector var names or truncated names
- `use_cache`: option to store hashed data locally to reuse offline

---
## *cancensus* provides tools to simplify working with the Census

```{r, eval=FALSE, message=FALSE, warning=FALSE, include=TRUE}
# List all Census variables
list_census_vectors(dataset)

# Search for Census variables containing specific terms 
search_census_vectors(searchterm, dataset)

# Search for parent and children nodes in Census variables
parent_census_vectors(vector)
child_census_vectors(vector)

# Provide detailed variable description for Census variable vectors
label_vectors(df)
```

---
## *cancensus* provides tools to simplify working with the Census

```{r, eval=FALSE, message=FALSE, warning=FALSE, include=TRUE}
# List all Census regions
list_census_regions(dataset)

# Search for Census regions containing specific terms 
search_census_regions(searchterm, dataset, level)

# Query for Census geography only without any data
get_census_geometry(dataset, level, geo_format)
```
---
## Calculating Language Diversity

The Language Diversity Index is calculated as 1 minus the probability any two random people in a population speak the same language. 

Greenberg (1958) provides a simple formula that we can use:
$$ 
LDI = 1 - \Sigma (P_i)^2
$$





Let's use this to find out what are the most diverse parts of Metro Vancouver.


---
### Calculating Language Diversity: Hands-on
```{r language_demo1, echo=TRUE, message=FALSE, warning=FALSE, results="hide"}
library(cancensus)

# Select Census dataset
dataset <- "CA16"

# search through Census variables to find the parent vector
language_total <- search_census_vectors("Language spoken most often", 
                                        dataset) %>% 
  filter(vector == "v_CA16_1355") 

# Select all leaf nodes of this vector. 
# The parameter TRUE returns only the finite leaves 
# among child nodes. 

language_children <- language_total %>%
  child_census_vectors(TRUE) 
# We'll need the aggregated total for our calculations 
# so let's append them together

language_vectors <- bind_rows(language_total, 
                              language_children) %>%
  pull(vector)

```
There are 218 language stems. 
---
### Calculating Language Diversity: Hands-on
```{r language_demo3, echo=TRUE, message=FALSE, warning=FALSE}
# Search for Census regions at the CMA level 
# with Vancouver in the name
van_region <- search_census_regions("vancouver", 
                                    dataset, 
                                    level = "CMA") %>%
  as_census_region_list 
```

---
### Calculating Language Diversity: Hands-on

Finally we can query our Census data combined with spatial data (in `sf` format*)
```{r language_demo5, echo=TRUE, message=FALSE, warning=FALSE}
library(sf)

langs_van <- get_census(dataset, level = "CSD", 
                        regions = van_region , 
                        vectors = language_vectors, 
                        geo_format = "sf", labels = "short",
                        quiet = TRUE)
```

.footnote[[*] cancensus can retrieve spatial data in either `sf` or `sp` format. The latter is a well developed and stable spatial library, while the former is bleeding-edge, often with bugs. Despite that `sf` is the future spatial standard for R and has a number of advantages.] 
---
### Calculating Language Diversity: Hands-on

We run the function* to calculate the Language Diversity Index on our Census data.
```{r language_demo_ldi, message=FALSE, warning=FALSE, include=FALSE}
ldi_calc <- function(df) {
  tidy_langs <- df %>%
  rename(`Language Total` = v_CA16_1355) %>%
  mutate(v_CA16_1364 = v_CA16_1364 + v_CA16_2153 + v_CA16_2150 + v_CA16_2159,
         v_CA16_1367 = v_CA16_1367 + v_CA16_2156 + v_CA16_2150 + v_CA16_2159) %>%
  select(-v_CA16_2153, -v_CA16_2150, -v_CA16_2159, -v_CA16_2156) %>%
  tidyr::gather(key = language, value = lang_count, v_CA16_1364:v_CA16_1937) %>%
  mutate(ldi_frac = ifelse(lang_count <= `Language Total`, lang_count/`Language Total`, 1)) %>%
  group_by(GeoUID) %>%
  mutate(ldi = 1 - sum((ldi_frac)^2)) %>%
  ungroup() %>%
  select(-language, -lang_count, -ldi_frac) %>%
  distinct()
  return(tidy_langs)
}
```
```{r language_demo6, echo=TRUE, message=FALSE, warning=FALSE}
van_ldi <- ldi_calc(langs_van)
```

.footnote[[*] The code to calculate the Language Diversity Index score for each geographic area is a bit verbose. You can find it detailed [here](https://www.dshkol.com/2017/language-diversity-in-canada/) along with a detailed description of some of the issues with language data if you are so inclined.]

---
### Calculating Language Diversity: Hands-on

Visualizing our data is very easy because our data is already in a nice and tidy format. 

```{r language_demo7, eval=FALSE, echo=TRUE, message=FALSE, warning=FALSE, fig.height=10, fig.width=16, dpi=300,  fig.align='center'}
library(ggplot2)

ggplot(van_ldi, aes(x = `Region Name`, y = ldi)) + 
  geom_bar(stat = "identity") + 
  coord_flip() + 
  theme_minimal() +
  labs(title = "Language Diversity for Greater Vancouver CSDs",
       x = NULL, y = "Language Diversity Index")
```

---
### Calculating Language Diversity: Hands-on

```{r language_demo8, echo=FALSE, message=FALSE, warning=FALSE, fig.height=10, fig.width=16, dpi=300,  fig.align='center'}
library(ggplot2)

ggplot(van_ldi %>% filter(ldi > 0) , 
       aes(x = `Region Name`, y = ldi)) + 
  geom_bar(stat = "identity") + 
  coord_flip() + 
  theme_minimal() +
  labs(title = "Language Diversity for Greater Vancouver CSDs",
       x = NULL, y = "Language Diversity Index",
       caption = "2016 Census data
       Some CSDs with very low populations excluded.")
```

---
### Calculating Language Diversity: Hands-on

We can easily compare population against language diversity.

```{r language_demo9, eval=FALSE, fig.height=10, fig.width=16, dpi=300,  fig.align='center', message=FALSE, warning=FALSE}
library(ggplot2)

ggplot(van_ldi %>% filter(ldi > 0) , aes(y = Population, x = ldi)) + 
  geom_point() + 
  theme_minimal() +
  labs(title = "Language Diversity vs Population Size",
       y = "Population Size", x = "Language Diversity Index",
       caption = "2016 Census data, 
       Some CSDs with very low populations excluded.")
```


---
### Calculating Language Diversity: Hands-on

```{r language_demo10, echo=FALSE, message=FALSE, warning=FALSE, fig.height=10, fig.width=16, dpi=300,  fig.align='center'}
library(ggplot2)
library(ggrepel)

van_ldi <- van_ldi %>% mutate(`Region Name` = as.factor(gsub(" \\(.*\\)","",`Region Name`)))

labeled_points <- c("Richmond", "White Rock","Burnaby", "Richmond","Vancouver","Coquitlam", "Surrey")
label_data <- van_ldi %>% filter(`Region Name` %in% labeled_points)

ggplot(van_ldi %>% filter(ldi > 0) , aes(y = log(Population), x = ldi)) + 
  geom_point() + 
  geom_label_repel(data = label_data,
    aes(y = log(Population), x = ldi, label = `Region Name`),
    size = 5,
    color = 'white', fill = "darkgrey",
    box.padding = unit(0.5, "lines"),
    point.padding = unit(0.75, "lines"),
    segment.color = 'grey50') + 
  theme_minimal() +
  labs(title = "Language Diversity vs Population Size for Greater Vancouver CSDs",
       y = "Log Population Size", x = "Language Diversity Index",
       caption = "2016 Census data, some CSDs with very low populations excluded.")
```
---
### Calculating Language Diversity: Hands-on

Finally, because our data is already spatial, making maps is super easy. 

```{r language_demo11, eval=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
library(sf)

ggplot(van_ldi) + 
  geom_sf(aes(fill = ldi), colour = NA) + 
  scale_fill_viridis_c("Language Diversity Index", option = "magma") + 
  theme_void() +
  labs(title = "Language Diversity Greater Vancouver CSDs",
       x = NULL, y = NULL,
       caption = "2016 Census data, 
       some CSDs with very low populations excluded.")
```
---
### Calculating Language Diversity: Hands-on
```{r language_demo12, echo=FALSE, message=FALSE, warning=FALSE, fig.height=10, fig.width=16, dpi=300,  fig.align='center'}
library(ggplot2)
library(sf)

map_theme <- theme_void() + 
  theme(plot.title=element_text(face="bold", hjust = 0.5)) + 
  theme(plot.subtitle=element_text(hjust = 0.5)) + 
  theme(plot.caption=element_text(size=8, margin=margin(t=10), hjust = 0.95)) +
  theme(legend.position = "bottom")

# Generate maps
ggplot(van_ldi) + geom_sf(aes(fill = ldi), colour = NA) + scale_fill_viridis_c("Language Diversity Index",option = "magma") +
  map_theme + 
  guides(fill = guide_colourbar(title.position="top")) +
  labs(title = "Language Diversity Greater Vancouver CSDs",
       x = NULL, y = NULL)
```
---
### Calculating Language Diversity: Hands-on

What if we used CT-level data instead? Easy. 

Just go back to the original cancensus query and swap the aggregation level. You already have the code for everything else. 

```{r language_demo13, message=FALSE, warning=FALSE, results = 'hide'}
langs_van <- get_census(dataset, level = "CT", 
                        regions = van_region , 
                        vectors = language_vectors, 
                        geo_format = "sf", labels = "short",
                        quiet = TRUE)
```
All you have to do now is rerun `van_ldi <- ldi_calc(langs_van)` on the new data and you're ready to go. 
---
### Calculating Language Diversity: Hands-on
```{r language_demo14, echo=FALSE, message=FALSE, warning=FALSE, fig.height=10, fig.width=16, dpi=300,  fig.align='center'}
library(ggplot2)
library(sf)

van_ldi <- ldi_calc(langs_van)

map_theme <- theme_void() + 
  theme(plot.title=element_text(face="bold", hjust = 0.5)) + 
  theme(plot.subtitle=element_text(hjust = 0.5)) + 
  theme(plot.caption=element_text(size=8, margin=margin(t=10), hjust = 0.95)) +
  theme(legend.position = "bottom")

# Generate maps
ggplot(van_ldi) + geom_sf(aes(fill = ldi), colour = NA) + scale_fill_viridis_c("Language Diversity Index",option = "magma") +
  map_theme + 
  guides(fill = guide_colourbar(title.position="top")) +
  labs(title = "Language Diversity Greater Vancouver CSDs",
       x = NULL, y = NULL)
```
---
### Calculating Language Diversity: Hands-on

DA-level map? Sure, why not*.

```{r language_demo15, eval = FALSE, message=FALSE, warning=FALSE, results = 'hide'}
langs_van <- get_census(dataset, level = "DA", 
                        regions = van_region , 
                        vectors = language_vectors, 
                        geo_format = "sf", labels = "short",
                        quiet = TRUE)
```

.footnote[[*] Actually you may want to watch out with DA level data as it can take a fair bit of time to query, download, process, and map - and may burn through your API quota.]

---
### Calculating Language Diversity: Hands-on
```{r language_demo16, echo=FALSE, message=FALSE, warning=FALSE, fig.height=10, fig.width=16, dpi=300,  fig.align='center'}
mv1
```

---
## *cancensus*: an R interface to Canadian Census data

Developed by [Jens von Bergmann](https://twitter.com/vb_jens), [Aaron Jacobs](https://github.com/atheriel), and [Dmitry Shkolnik](https://twitter.com/dshkol)

Presented by Dmitry Shkolnik tw: @dshkol

Additional information, examples, and vignettes available on the github page for _cancensus_ at [github.com/mountainMath/cancensus](https://github.com/mountainmath.cancensus)

In depth examples can be found at Jens' blog [(doodles.mountainmath.com)](https://doodles.mountainmath.ca) or at [(dshkol.com)](https://www.dshkol.com). 

